### 2012-01-22 paper_draft.tex
Added a new introduction, structured the results and discussion, moved a lot of paragraphs

Genome-Wide Association Studies (GWAS) are a comparatively easy and cheap way of finding Single Nucleotide Polymorphisms (SNPs) of medical relevance. Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits. Genome-Wide Association Studies make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This method does not allow inference of causal differences but merely identifies correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

Although companies like 23andMe are willing to contribute to science, it is not easy for individual scientists to access the data. This is mainly due to concerns about privacy, liability and consent. Nevertheless, there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}. While this makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public. While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

A possible solution to this is a community-driven platform that aggregates genetic and phenotypic information of people who are willing to share their data with the general public and have given their informed consent. We designed an online survey to assess interest in such a crowdsourcing platform, in which we asked an audience interested in biology topics about their willingness to share their genetic and phenotypic information with the public. The aim of the survey was not to determine the percentage of the population interested in sharing genotypic information, but to get an idea of whether a crowdsourcing platform could get enough support to go forward with the project. Additionally, we built a platform which allows customers of DTC genetic testing to publicize genetic and phenotypic information and gives researchers multiple ways of reusing the data.

@@

\subsection*{openSNP}
We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages.

The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories. Every user can add completely new phenotypes that are not yet queried by openSNP. To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database.

OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.

Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files. The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs. In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes. See figure n for a distribution of data acquisition over time.

The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype provided by the most users is the eye color, which has been entered by 54 users. There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.

In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.

The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

\section*{Discussion}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation. Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods. Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves. As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.

@@

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.

@@

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Survey on Sharing Genetic Information}
The survey was done with Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook. Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.

>>>>

Genome-Wide Association Studies (GWAS) are a comparatively easy and cheap way of finding Single Nucleotide Polymorphisms (SNPs) of medical relevance. Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits. Genome-Wide Association Studies make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This method does not allow inference of causal differences but merely identifies correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
\\
\\
FZ REWRITE 
\\
The comparison of DNA sequences from indivduals in a population reveals sites that are variable. 
These variations are called Single Nucleotide Polymorphisms (SNP). Each different nucleotide 
at a specific site is defined as an allele. FZ: Explain genotype!!
Different methods have been developed to measure/find/access/sequence SNPs in an individual, a
process called genotyping. Microarrays are a cheap and fast way to genotype individuals, in comparison
to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associated SNPs with for example diseass. Genome-Wide Assocation Studies
(GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
enables GWAS to find significant overrepresantations of alleles in patients and associate them wir a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

Although companies like 23andMe are willing to contribute to science, it is not easy for individual scientists to access the data. This is mainly due to concerns about privacy, liability and consent. Nevertheless, there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.

FZ: The data is scattered and unorganized, making it hard to use this infornomation. While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

We designed an online survey to evaluate the support in the community for a crowdsourced online plattform. Here we present the results of the survey in combination with the openSNP project. The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. The data is made available to research via a Creative-Commones-Zero licence.

@@

\subsection*{openSNP}

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.

\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files. The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
\subsubsection*{Crowdsourcing phenotypes}
Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages. The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories. Every user can add completely new phenotypes that are not yet queried by openSNP. To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database. In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes. See figure n for a distribution of data acquisition over time.
The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype provided by the most users is the eye color, which has been entered by 54 users. There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.  
\subsubsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.

\section*{Discussion}

\subsubsection*{Survey issues}
Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
\subsubsection*{GWAS and open data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation. Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods. Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves.

@@

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
\subsubsection*{Why should a user publish their genetic informationß Pro vs. Con}
With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising. 
\subsubsection*{Privacy and health implications and ethical considerations}

As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

@@

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Survey on Sharing Genetic Information}
The survey was done with Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
